---
title: "L08 - Wrangling and (more) Data Verbs"
subtitle: "Data Computing Chapter 10"
author: |
        | Presenter: Olivia Beck
        | Content credit: Matthew Beckman
date: "May 31, 2023"
output: 
  slidy_presentation: default
  html_notebook: default
---

```{r include=FALSE}
# Frontmatter
rm(list = ls())

# packages
library(mosaicData)
library(knitr)
library(tidyverse)
library(dcData)

```


## Agenda

- Basic Info about the Midterm and Final Project
- Discussion of common issues/questions 
- Review of data verbs & summary functions with your small group
- Discussion of new functions and key concepts
- Guided example highlighting several new data verbs
- RStudio Concept Maps: <https://github.com/rstudio/concept-maps>


## Midterm and Final Project

### Midterm 

- June 9 (Friday) during class 
- On paper 
- Anything we have done in class before then is fair game for questions
- This is not going to be a difficult exam. It will cover material from `tidyverse` functions and GitHub workflows. 


To practice: 
- Look through textbook and read through some of the problems not assigned 
- Make sure you understand everything from the activities, homeworks, and reading quizzes
- If you have questions, ask! Post general content question on Canvas Discussion board so your peers can learn too
- I will provide some practice questions soon 



### Final Project 

- Multiple due dates (on canvas)
- Goal is to do a "professional" data analysis that is REPRODUCIBLE 
  - Research question
  - Data Exploration
  - Answer your question with wrangled data and graphs 
    - NO STATISTICAL ANALYSIS
- You need to find your own data sets (minimum of 2)




## Three Important Concepts (Again)

1. Data can be usefully organized into tables with "cases" and "variables."  With "tidy data"
    - every row corresponds to a distinct case (e.g. a person, a car, a year, a country in a year) 
    - columns represent variables/features of the cases

2. Data graphics and "glyph-ready" data
    - each case corresponds to a "glyph" (mark) on the graph
    - each variable to a graphical attribute of that glyph such as x- or y-position, color, size, length, shape, etc. 
    - similar story applies for modeling purposes

3. When data are not yet in glyph-ready form, you can transform (i.e. wrangle) them into glyph-ready form.  
    - Such transformations are accomplished by performing one or more of a small set of basic operations on data tables
    - This is the work of "data verbs" and other functions



## `select`

The use of select() is straightforward. It takes a data frame as input along with the names of one or more variables and outputs the data frame with only those variables. 

```{r}
data(CountryData) #dcData package 

## list all the column names
names(CountryData)

```



### Only look at Country, Population, and Area

```{r}
Tab1 <- CountryData %>% 
  select(country, pop, area)

head(Tab1)

```


## `rename`

`rename` takes in a variable name, and renames that variable. We use `rename(NewName = OldName)`

### rename "area" to "kmSquared"

```{r}
head(Tab1)

Tab2 <- 
  Tab1 %>%
  rename(kmSquared = area)

head(Tab2)

```

## `filter`

To filter means to remove unwanted material. The data verb `filter()` removes unwanted cases, passing through to the result only those cases that are wanted or needed.


With filtering, the cases to pass through are specified by one or more criteria or tests. The tests usually involve comparisons between variables or between a variable and a scalar.


Filtering and selecting are complementary actions for subsetting a data frame. `select()` passes the specified subset of *variables*; `filter()` passes the specified subset of *cases*.


```{r}
data(BabyNames)

head(BabyNames)

```

### display only female names 

```{r}
# Method 1
Method1 <- 
  BabyNames %>%
  filter(sex == "F")

head(Method1)

# Method 2
Method2 <- 
  BabyNames %>%
  filter(sex != "M")

head(Method2)

```


### Display only babys born after 1990

```{r}
Baby1990 <-
  BabyNames %>%
  filter(year > 1990)

head(BabyNames)
```

### Display only female babies born after 1990

```{r}
F1990 <- 
  BabyNames %>%
  filter(sex == "F",
         year > 1990)

head(F1990)
```

### Display babies that are female OR born after 1990

```{r}
Baby1990F <-
  BabyNames %>%
  filter(sex == "F" | year > 1990)

sample_n(Baby1990F, 10)


```


## `mutate`

The word “mutate” means to change in form or nature. The data verb `mutate()` is more specific: to change a variable or add new variables based on the existing ones. Mutation leaves the cases exactly as they were; the output of `mutate()` will have exactly as many rows as the input data frame


### Calculate the population density 

```{r}
head(Tab1)

PopDens <- 
  Tab1 %>%
  mutate(pop_dens = pop / area)

head(PopDens)

```

### Caution! 

If you use a old vairable name for a new varaible in a `mutate` statement, it will overwrite the old variable.

```{r}
head(Tab1)

Tab1.Overwrite <-
  Tab1 %>%
  mutate(area = 2 * area)

head(Tab1.Overwrite)


```




## `arrange`

`arrange` will arrange the rows by the variable you provide. 

```{r}
head(Tab1)

Tab1 %>%
  arrange(pop) %>% 
  head()

Tab1 %>%
  arrange(desc(pop)) %>%
  head()

```


## Presidential Investigation 

- Let's take some of our new tools for a spin...
- The `presidents` data are available from the `tidyverse` package (actually `ggplot2` which was loaded by `tidyverse`)
- we'll add a few other presidents to the data as well

```{r echo=TRUE}
library(tidyverse)
library(dcData)
library(tibble)     # we'll use `tibble::tribble()`  to specify new rows 
library(lubridate)  # nice functions for handling dates

# data intake
data("presidential", package = "ggplot2")

## Add some more cases to the data set 
# you can search the help for the `tribble` function if you want to figure it out what's happening
AddPres <- 
    tibble::tribble(~name, ~start, ~end, ~party,
                     "Roosevelt", ymd("1933-03-04"), ymd("1945-04-12"), "Democratic",
                    "Truman", ymd("1945-04-12"), ymd("1953-01-20"), "Democratic",
                    "Trump", ymd("2017-01-20"), ymd("2021-01-20"), "Republican", 
                    "Biden", ymd("2021-01-20"), ymd(today()), "Democratic")

# append the rows of our two "data frames"
Presidents <- bind_rows(presidential, AddPres)

head(Presidents)
```

### Display only baby names from 1980, 1990, 2000, or 2010

```{r}
BabyDecade <- 
  BabyNames %>% 
  filter( year %in% c(1980, 1990, 2000, 2010))

BabyDecade %>%
  sample_n(10)

```





## Guided practice 

1. various summary/exploration functions to meet the Presidents
2. calculate `duration` of presidency 
3. show all records for presidents named "Bush"
4. subset the data to include only `name`, `party`, `duration`
5. which 3 presidents served shortest?
6. which 3 presidents served longest?
7. which presidents served more than 4 years, but less than 8 years?
8. investigating the power balance
   - which party has most presidents?
   - which party spent most days in power?
9. add Harry Truman & Franklin D. Roosevelt, then repeat 6-9


## Answers 

```{r}
# 1. summary functions
glimpse(Presidents)  # what's the benefit?
str(Presidents)      # ?
summary(Presidents)  # not so useful?!
#View(Presidents)     # console please!


# some wrangling 
Presidents_wrangled <- 
    Presidents %>%
    mutate(duration = end - start,                                     # 2
           name = if_else(name == "Bush" & start < ymd("2000-01-01"),  # 3
                          true = "Bush (Sr)", 
                          false = name)) %>%
    select(name, party, duration)                                      # 4
    

### some results
# 3 check the bushes
tail(Presidents_wrangled) 

# 6 shortest
Presidents_wrangled %>%
    arrange(duration) %>%
    head()

# 7. longest  
Presidents_wrangled %>%
    arrange(duration) %>%
    tail()

Presidents_wrangled %>%
    arrange(desc(duration)) %>%
    head()

# 8. more than 4 years, but less than 8 years

Presidents_wrangled %>%
    filter(duration > 1461, duration < 2922)


# 9. party balance

Presidents_wrangled %>%
    group_by(party) %>%
    summarise(total = n(), 
              in_power = sum(duration))

# repeat!

```

