---
title: "Redefining rows: Wide vs narrow data organization"
subtitle: "Data Computing Chapter 12"
author: |
        | Presenter: Olivia Beck
        | Content credit: Matthew Beckman
output: 
  slidy_presentation: default
  html_notebook: default
---

```{r include=FALSE}
library(tidyverse)
library(printr)
library(dcData)

#devtools::install_github("dcl-docs/dcldata")
library(dcldata)

options(width = 80)
```

## Cases, Variables, and Values

A data table is comprised of *cases* and *variables*.

Each *variable* comprises *values* (or levels).

There is no hard distinction between a variable and a value.  What's a variable in one situation may be a value in another, and vice versa.

Here are 2 tables with the same information. 

Q: What IS different?
Q: What is NOT different?

Table 1: 
```{r echo=FALSE}
example_eagle_nests
```


Table 2: 

```{r echo=FALSE}
example_eagle_nests_tidy
```


## Cases, Variables, and Values

#### For Table 1 .... 

- Variables: `2007`, `2009`
    - Values: 
        - `2007`: count of nests
        - `2009`: count of nests
- Cases: Regions {Pacific, Southwest, Rocky Mountains and Plains}

```{r echo=FALSE}
example_eagle_nests
```


#### For Table 2

- Variables: `year`, `num_nests`
    - Values: 
        - `year`: 2007 and 2009
        - `num_nests`: count of nests
- Cases: Regions and Year

```{r echo=FALSE}
example_eagle_nests_tidy
```

Neither of the tables are wrong, but sometimes one form is more helpful than the other. 


## Two formats

- Data in Key/Value format are **narrow**
    - possible to get *too* narrow if the meaning of case becomes awkward
    
| ID      | Key | Value |
| ----------- | ----------- | ----------- |
| ID1      | Key1       | value(1,1)      |
| ID1      | Key2       | value(1,2)      |
| ... | ... | ... | 
| ID2      | Key1       | value(2,1)      |
| ID2      | Key2       | value(2,2)     |
| ... | ... | ... | 

- The corresponding **wide** format has 
    - separate variables for each level in `key`
    - sets the values for those variables from the info in `value`
    
    
| ID      | Key1 | Key2 | ....|
| ----------- | ----------- | ----------- | ---|
| ID1      | value(1,1)      | value(1,2)       |....|
| ID2      | value(2,1)      | value(2,2)       |....|
| ... | ... | ... | 


### Narrow


Let's remember our original table: 

```{r}
example_eagle_nests
```


Let's make it narrow

```{r}
narrow_table <- 
  example_eagle_nests %>% 
  pivot_longer(
    cols = c(`2007`, `2009`),  #names of columns we want to become a new variable 
    names_to = "year",         #what you want to call the new column of the data in the line above
    values_to = "num_nests"    #what you want to call the variable that stores the values
  )

narrow_table
```


This is a good narrow table (and a tidy table!). Each case (region, year, quarter combination) has a row and each variable has a column. 

Be careful you don't make your tables too narrow where we loose the definition of a case (example on next slide). 


### Wide


Let's remember our narrow table. 

```{r}
narrow_table

```


Now we can make our narrow table wide again. 

```{r}
wide_table <- 
  narrow_table %>%
  pivot_wider(
    names_from = year,       # the column in the narrow table with the column names for the wide table 
    values_from = num_nests  # the column in the narrow table with the values in the narrow wide
  )

wide_table
```

We can compare this to the original table to make sure its the same thing. Yep! 

```{r}
example_eagle_nests
```



## (Too) Narrow
 

```{r echo=FALSE, warning=FALSE, echo = F}
###### Data set up, you don't need to know this 
quarters <- rpois(4, 2)+1
quarters <- quarters/sum(quarters)

class_example_Table <- 
  example_eagle_nests %>% 
  pivot_longer(
    cols = c(`2007`, `2009`),
    names_to = "year",         
    values_to = "num_nests"    
  ) %>%
  slice(rep(1:n(), each = 4))%>%
  mutate(quarter = rep(quarters, 6)) %>%
  mutate(ByQuarter = round(num_nests * quarter)) %>%
  mutate(Quarter = rep(c("Q1", "Q2", "Q3", "Q4"), 6)) %>%
  select(region, year, Quarter, ByQuarter) %>%
  pivot_wider(names_from = c(year, Quarter), values_from = ByQuarter )
```


Say this is the data set we have: 

```{r}
class_example_Table

```

We can make it (appropriately) narrow like this: 

```{r}
class_example_Table %>%
  pivot_longer(!region,                         #everything but region
               names_to = c("Year", "Quarter"), 
               names_sep = "_",                  #how year and quarter are seperated 
               values_to = "NestCount")

```


But we could have made it too narrow. None of the data has been lost here, but it's not a helpful form since there isn't a useful definition of "case". 

```{r}
names <- colnames(class_example_Table)

class_example_Table %>%
  mutate_all(as.character) %>%
  pivot_longer(cols = names,   
               names_to = "key", 
               values_to = "value") 

```




## So what?

- This allows us to easily redefine how rows are presented in the data
    - Possibly motivated by the research question
    - Possibly motivated by desire to join two data tables with different case definitions
    - Possibly motivated by a data visualization
- Also, some operations are easy in wide format, but hard in narrow and *vice versa*
- We need tools that make it easy to switch back and forth


## Example from `BabyNames`

```{r echo=FALSE}
data("BabyNames", package = "dcData")

ShortBabyNames <- 
  BabyNames %>%
  filter( name %in% c("Olivia", "Zoe", "Quentin"), 
          year %in% 2012:2013 ) %>% 
  arrange(name, year)
```

```{r echo=FALSE}
ShortBabyNames
```

### Questions:

RQ 1. How many babies of each name and sex?
RQ 2. For each name, is it primarily given to girls or boys?  Which names are gender neutral?



## In narrow format

```{r}
data("BabyNames", package = "dcData")

BabyNames <- 
  BabyNames %>%
  filter( name %in% c("Leslie", "Morgan", "Blake", "Taylor")) 
```

RQ 1. How many babies of each name and sex?

```{r}
BabyTotals <-
  BabyNames %>%
  group_by(name, sex) %>%
  summarise(total = sum(count))
```

```{r echo=FALSE}
BabyTotals
```

Easy!



## In Wide format

RQ 2. Which names are most gender neutral?


```
WideOutput <- 
  NarrowInput %>% 
  pivot_wider(names_from = var1, values_from = var2, values_fill = 0)
```

- we want a new column for each category of `sex`, so `names_from = sex` 
    - we will "unstack" each available category as a new variable (a.k.a. cast, spread, unfold)
    - categories of `sex` were "F" and "M" in this example
- the values/entries for our new variables are coming from `total`, so `values_from = total`
    - `values_fill = 0` specifies a default value to fill when missing


```{r}
BabyTotalsWide <- 
  BabyTotals %>% 
  pivot_wider(names_from = sex, values_from = total, values_fill = 0)

BabyTotalsWide
```

## With sexes side by side... 

We can easily calculate balance associated with names


```{r}
BabyTotalsWide <- 
  BabyTotalsWide %>% 
  rename(fem = F, male = M) %>%         # `F` is a terrible variable name (why?)
  mutate(prop_fem  = fem  / (male + fem), 
         prop_male = male / (male + fem),
         name_specificity = pmax(prop_fem, prop_male))    # what does `pmax()` do?

BabyTotalsWide
```

## `pivot_longer( )`---when you have "Wide" and want "Narrow"

Syntax:

```
NarrowOutput <- 
  WideInput %>% 
  pivot_longer(cols = c(wide_var1, wide_var2, ...), names_to = "long_var1", values_to "long_var2")
```

- The `cols` are the variables we want to combine (a.k.a. melt, stack, fold, gather)
- e.g. `prop_fem` and `prop_male` in this case

```{r echo=TRUE}
BabyTotalsNarrow <- 
  BabyTotalsWide %>% 
  select(prop_fem, prop_male) %>%
  pivot_longer(cols = c(prop_fem, prop_male), names_to = "sex", values_to = "proportion") 

BabyTotalsNarrow
```

## With sexes stacked again...

We can make an intuitive bar chart (though some clean up is needed...)

```{r}
BabyTotalsNarrow %>%
  ggplot() + 
  geom_bar(aes(x = name, fill = sex, weight = proportion)) 
  
```

#### With some improvements 

- clean up labels of sexes
- add title, source, & better axis labels (y-axis label had been flat wrong)

```{r}
# first, clean up the labels in `sex` for plotting
BabyTotalsNarrow %>%
  mutate(sex = if_else(sex == "prop_fem", 
                       true = "female", 
                       false = if_else(sex == "prop_male", 
                                       true = "male", 
                                       false = "unk")  # end of "inner" if_else()
                       )                               # ends the "outer" if_else()
         ) %>%                                         # ends the mutate() 
  ggplot() + 
  geom_bar(aes(x = name, fill = sex, weight = proportion)) + 
  ggtitle("Gender Balance among Names of Beckman Kids", 
          subtitle = "source: U.S. Social Security Administration") + 
  xlab("Name") + 
  ylab("Proportion")
  
```
