---
title: "Regular Expressions (RegEx)"
subtitle: "Data Computing Ch 17"
author: |
        | Presenter: Olivia Beck
        | Content credit: Matthew Beckman
output: 
    slidy_presentation: default
    html_notebook: default
---




```{r include=FALSE}
require(lubridate)
require(tidyverse)
require(rvest)
knitr::opts_chunk$set(message = FALSE)
```


## Chapter 17: Key Ideas

* Regular expressions allow us to match meaningful **patterns** in character strings
* Some popular uses:
    * detect whether a **pattern** is contained in a string (use `filter()` & `grepl()`)
    * substitute the elements of that **pattern** with something else (use `mutate()` & `gsub()`)
    * extract a component that matches the **pattern** (use `tidyr::extract()`)
    
    
  


## Some Exploits in the Land of RegEx 

* Medtronic, Inc - quality monitoring for medical technology
    * Match key word or phrase in offline complaint data (uncommon)
    * Subset of complaint data and evaluate rate of some outcome over time

* PSU Men's Volleyball
    * Teams now have access to complete data for play in every match
    * Using RegEx to help parse the data to gain competitive advantage for PSU
    * (Sort of like Moneyball for Volleyball...)

* Scraping HTML data
    * We scraped the [Men's Pole Vault World Records from Wikipedia](https://en.wikipedia.org/wiki/Men%27s_pole_vault_world_record_progression)
    * We had been stuck because of footnotes in the `Date` column previously
    * We can use RegEx to clean up (We'll try it together in a few moments)


## How to Survive in the Land of Regex

* Step 1: Memorize the following special characters and their use: \\d, \\w, \\S, [0-9], [^0-9], [[:lower:]], [[:alnum:]], \\W, \n, ?, ., $, %, |, \\<, ^, \\, {3}, *, +, \\s, \\B, \\>, \\x

## How to Survive in the Land of Regex

* **NO!!!** absolutely no need to memorize all of it
* Use the RStudio Cheat Sheet: <https://www.rstudio.com/resources/cheatsheets/>
* Use Google
* Just like everything else in (R) Programming:
    * Don't start from scratch
    * Find working code that does something similar
    * Make many iterations of small changes checking at each change that it didn't break
    * Keep going until the original code evolves into the thing you want!
    
Here are some common RegEx expressions 

* A single . means “any character.”
* A character, e.g., b, means just that character.
* Characters enclosed in square brackets, e.g., [aeiou] means any one of those characters. (So, [aeiou] is a pattern describing a vowel.)
* A vertical bar means “either.” For instance, the regex "a|e|i|o|u" will match any vowel.
* The ^ inside square brackets means “any except these.” So, a consonant is [^aeiou]
* ^ at the beginning of a regular expression means “the start of the string”
* $ at the end means “the end of the string.”
* Two simple patterns in a row, means those patterns consecutively. Example: "M[aeiou]" means a capital M followed by a single lower-case vowel.
* A pattern followed by a + means “zero or more times.” So, "M[aeiou]+" matches “Miocene”, “Mr.”, “Ms”, “Mrs.”, “Miss”, and “Man”, among many others.
* A simple pattern followed by a ? means “zero or one times.” So, "M[aeiou]?[^aeiou]" will not match “Miocene”. (The regex means: M followed by at most one vowel, followed by a consonant. “Miocene” has two vowels following the M.)
* A simple pattern followed by a * means “one or more times.” "M[aeiou]*" will match “Miocene” and “Miss”, but not “Mr.”
* A simple pattern followed by {2} means “exactly two times.” Similarly, {2,5} means between two and five times, {6,} means six times or more. For instance, [aeiou]{2} means “exactly two vowels in a row” and will not match “Miss” or “Mr” or “Madam”.

  
## Detection

We can use `filter(grepl(pattern, object_to_search))` to detect if a pattern is present 

Example: Find all baby names that contain "shine" 

```{r}
library(dcData)
data(BabyNames)

#first make a list of all the unique names
UniqueNames <- 
  BabyNames %>%
  select(name) %>%
  unique() %>%
  arrange(name)

head(UniqueNames, 10)

UniqueNames%>% 
  filter(grepl("shine", name, ignore.case = TRUE))
```

Example: Find the names that contain three or more vowels in a row

```{r}
UniqueNames %>% 
  filter(grepl("[aeiou]{3,}", name, ignore.case = TRUE))  %>%
  sample_n(10)

```

Example: Find the names that contain three or more consonants in a row

```{r}
UniqueNames %>% 
  filter(grepl("[^aeiou]{3,}", name, ignore.case = TRUE)) %>%
  sample_n(10)

```


## Substitution/ Replacement

We use `mutate(new_variable = gsub(pattern, replacement, old_variable))` to find a pattern and replace it with a new one. 

Say I have the following table about public debt by country: 

```{r}
Debt <- data.frame(
  country = c("United States*", "Japan", "China", "Germany", "Italy"),
  debt = c("$17,607 billion", "$9,872 billion", "$3,894 billion", "$2,592 billion", "$2,334 billion"), 
  percGDP = c("73.60%", "214.30%", "31.70%", "81.70%", "126.10%")
)

Debt

```
To use these numbers for computations, they must be cleaned up.

```{r}
Debt %>% 
  mutate(debt = gsub(pattern = "[$,]|billion", replacement = "", debt),
         percGDP = gsub(pattern = "[%,]", replacement = "", percGDP))

```
#### Remove a currency sign

```{r}
currency <- c("$100.95", "45¢")
gsub(pattern = "\\$|€|¥|£|¢$", replacement = "", currency)
```

#### Remove leading or trailing spaces

```{r}
string <- "   My name is Julia     "
gsub(pattern = "^ +| +$", replacement = "", string)
```



## Extraction 

Use `muatate(new_variable = str_extract(old_variable, pattern))` from to extract a component that matches the patterns.

Example: Find baby names with 4 consecutive vowels, and tell me what those vowels are. 

```{r}
UniqueNames %>%
  filter(grepl("[aeiou]{4,}", name, ignore.case = TRUE))%>%
  mutate(match = str_extract(string = name, pattern = "([aeiouAEIOU]{4})"))%>%
  sample_n(10)

```


## Example: Pole Vault Records Clean Up 



```{r include=FALSE}
require(mosaic)
require(lubridate)
require(tidyverse)
require(rvest)
knitr::opts_chunk$set(message = FALSE)
```


```{r echo=TRUE}
webpage <- "https://en.wikipedia.org/wiki/Men%27s_pole_vault_world_record_progression"

table_list <- 
  webpage %>%
  read_html(header=TRUE) %>%
  html_nodes(css = "table") %>%
  html_table(fill = TRUE)

PVRecords <- table_list[[2]]  # convert list to data frame
head(PVRecords, 3) # inspect the data

```


#### Tasks to clean up: 

1. we should fix the variable name representing the number of world records achieved by each athlete
2. locate and replace all footnotes in the `Date` variable using `gsub()`
3. convert `Date` to a date class variable in R using a `lubridate` function
4. use `tidyr::extract()` to store the metric heights from the `Record` variable (make sure there are no spaces )


## Solutions

```{r}

# locate and replace all footnotes in the `Date` column
PVMen <- 
  PVRecords %>%
  rename(recordsBroken = `#[4]`) %>%
  mutate(Date = gsub(pattern = "\\[.\\]", replacement = "", x = Date)) %>%
  mutate(Date = lubridate::mdy(Date)) %>% #convert to date
  tidyr::extract(col = Mark, into = "Meters", regex = "(^\\d\\.\\d)") %>%
  mutate(Meters = parse_number(Meters)) #convert to numeric(drops non-numeric characters)
  
PVMen %>% 
  head()
```




## Cool Graphs 

```{r}
PVMen %>%
  ggplot(aes(x = Date, y = Meters)) + 
  geom_step() + 
  geom_point(alpha = 0.5, aes(color = Nation))+
  scale_color_brewer(palette = "Spectral")

```

## What happened?


Here's a video explanation if you want to watch the evolution: 

<https://youtu.be/QGTdEhUW0nE>

